Pipit on the Post: Proving Pre- and Post-conditions of Reactive Systems

### 2024-01-16 2024pipit/section/20-Motivation.tex
eval: kind2
copy

To actually prove the property above, we use the meta-language \fstar{}'s tactics to translate the program into a transition system and prove the property inductively on the system.
Finally, we \emph{bless} the expression, which marks the properties as valid ($[\PSUnknown := \PSValid]$).
Blessing is an intensional operation: it traverses the expression and updates the internal metadata, but it does not affect the runtime semantics.

@@

We impose three main restrictions on the triggers array: the time-marks must be sorted; there must be an adequate time-gap between any two triggers that are enabled on the same cycle index; and each trigger's time-mark must be greater-than-or-equal to its index.

With these restrictions in place, we prove a lemma \emph{lemma_can_reach_next}, which states that for all valid cycle indices and trigger indices, if the current trigger is enabled in the current cycle and there is another enabled trigger scheduled to occur somewhere in the array after the current one, then there is an adequate time-gap to allow the driver to skip over any disabled triggers in-between.
These properties are straightforward in a theorem prover, but would be difficult to state in a model-checker with a limited specification language.

@@

\begin{tabbing}
  @MM@\= @MM@ \= @MMMMMMMMMMMMM@ \= \kill
  @let@ $\text{trigger_fetch}_{\PSValid}$ ($\textit{cycle}$: $\NN$): $\stream \NN \to \stream \NN$ = \\
  \> @let@ $\textit{contract} = \text{Contract.contract_of_stream}_1$ \{ \\
  \> \> @rely@ = $(\lambda \textit{time}.~ \text{time_no_skips}~\textit{time} )$ \\
  \> \> @guar@ = $(\lambda \textit{time index}.~ (\text{index_valid}~\textit{index} \wedge \text{enabled}~\textit{index}~\textit{cycle})$ \\
  \> \> \> $\implies (\text{time_mark}~\textit{index}) \ge \textit{now})$ \\
  \> \> @body@ = $(\lambda \textit{time}.~ \text{trigger_fetch}~\textit{cycle}~\textit{time} )$ \\
  \> \} @in@ \\
  \> @assert@ $(\text{Contract.inductive_check}~\textit{contract})$ @by@ (pipit_simplify ()); \\
  \> $\text{Contract.stream_of_contract}_1~\textit{contract}$
\end{tabbing}

>>>>

To actually prove the property above, we use the meta-language \fstar{}'s tactics to translate the program into a transition system and prove the property inductively on the system.
Finally, we \emph{bless} the expression, which marks the properties as valid ($[\PSUnknown := \PSValid]$).
Blessing is an intensional operation that traverses the expression and updates the internal metadata, but does not affect the runtime semantics.

@@

We impose three restrictions on the triggers array: the time-marks must be sorted; there must be an adequate time-gap between any two triggers that are enabled on the same cycle index; and each trigger's time-mark must be greater-than-or-equal to its index.

With these restrictions in place, we prove a lemma \emph{lemma_can_reach_next}, which states that for all valid cycle indices and trigger indices, if the current trigger is enabled in the current cycle and there is another enabled trigger scheduled to occur somewhere in the array after the current one, then there is an adequate time-gap to allow the driver to skip over any disabled triggers in-between.
These properties are straightforward in a theorem prover, but are difficult to state in a model-checker with a limited specification language.

@@

\begin{tabbing}
  @MM@\= @MM@ \= @MMMMMMMMMMMMM@ \= \kill
  @let@ $\text{trigger_fetch}_{\PSValid}$ ($\textit{cycle}$: $\NN$): $\stream \NN \to \stream \NN$ = \\
  \> @let@ $\textit{contract} = \text{Contract.contract_of_stream}_1$ \{ \\
  \> \> @rely@ = $(\lambda \textit{time}.~ \text{time_no_skips}~\textit{time} );$ \\
  \> \> @guar@ = $(\lambda \textit{time index}.~ (\text{index_valid}~\textit{index} \wedge \text{enabled}~\textit{index}~\textit{cycle})$ \\
  \> \> \> $\implies (\text{time_mark}~\textit{index}) \ge \textit{time});$ \\
  \> \> @body@ = $(\lambda \textit{time}.~ \text{trigger_fetch}~\textit{cycle}~\textit{time} );$ \\
  \> \} @in@ \\
  \> @assert@ $(\text{Contract.inductive_check}~\textit{contract})$ @by@ (pipit_simplify ()); \\
  \> $\text{Contract.stream_of_contract}_1~\textit{contract}$
\end{tabbing}

### 2024-01-16 2024pipit/section/30-Core.tex
copy

\autoref{f:core:structure} shows the high-level architecture of Pipit.
On the left-hand-side, the surface syntax embedded in \fstar{} is shown; this includes some Pipit-specific syntactic sugar.
The translation from the surface syntax to the core language is trusted.
There are two targets from the core language: abstract transition systems for verification, and executable transition systems for extraction to C.
The translation from the abstract systems is verified to be an abstraction according to the dynamic semantics (\autoref{s:core:dynamic}).
One property that still remains to be proven is that the proof obligations on the abstract transition system entail the original proof obligations (\autoref{s:transition}); this pending proof is denoted as negated entailment in the figure ($\not\vdash$).
The translation to executable transition system is proven to be semantics preserving, as is the subsequent translation to \lowstar{}.
The translation from \lowstar{} to C is external to this paper and forms part of our trusted computing base.

@@

Recursive streams, which can refer to previous values of the stream itself, are defined using the fixpoint operator ($\xrec{x}{e[x]}$); the syntax $e[x]$ means that the variable $x$ can occur in $e$.
% To ensure that streams are productive,
As in Lustre, recursive streams can only refer to their previous values and must be \emph{guarded} by a delay: the stream $(\xrec{x}{\xfby{0}{(x + 1)}})$ is well-defined, but stream $(\xrec{x}{x + 1})$ is invalid and has no computational interpretation.
This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
% We use this form to define a substitution-based operational semantics that is syntax-directed, as opposed to the mutually-recursive form in \cite{caspi1995functional} which is not syntax-directed.
% Our semantics has a simpler proof of determinism; we believe it has simplified other necessary proofs too and will perform further evaluation.
Although we cannot express mutually-recursive bindings in the core syntax here, we can express them as a notation on the surface syntax by combining the bindings together into a single tuple.

@@

Contracts $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ involve two verification conditions.
Firstly, when a contract is \emph{defined}, the definer must prove that the body $\ebody$ satisfies the contract: roughly, if $\erely$ is always true, then $\eguar[x := \ebody]$ is always true.
Secondly, when a contract is \emph{instantiated}, the caller must prove that the environment satisfies the precondition: that is, $\erely$ is always true.
Conceptually, then, a contract could have two property statuses: one for the definer, and one for the instantiation.
However, in practice, it is not useful to defer the proof of a contract definition --- one could achieve a similar effect by replacing the contract with its implementation.
For this reason, we only annotate contracts with one property status, which denotes whether the instantiation has been proved to satisfy the precondition.

@@

For values, we use an auxiliary judgment form $\mtypingval{v}{\tau}$ to denote that value $v$ has type $\tau$.
Likewise, for primitives we use the auxiliary judgment form $\mtypingprim{p}{(\tau_1 \times \cdots \hdots \times \tau_n) \to \tau'}$ to denote that primitive $p$ takes arguments of type $\tau_i$ and returns a result of type $\tau'$.
Primitives are pure, non-streaming functions.

@@

Finally, rule \textsc{TContract} applies for a contract $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ with a body expression of some type $\tau$.
The overall expression has result type $\tau$.
Both rely and guarantee clauses must be boolean expressions.
Additionally, the guarantee clause can refer to the result value by $x$.

@@

The dynamic semantics of Pipit are defined in \autoref{f:core-bigstep}.
We present our semantics in a big-step form.
This differs somewhat from traditional \emph{reactive} semantics of Lustre~\cite{caspi1995functional}.
Our big-step semantics emphasises the equational nature of Pipit, as it is substitution-based, while the reactive semantics emphasises the finite-state streaming execution of the system.
We use transition systems for reasoning about the finite-state execution (\autoref{s:transition}), which is fairly standard~\cite{brun2023equation,champion2016kind2,raymond2008synchronous}.
Previous work on the {\sc W-calculus}~\cite{gallego2021w} for linear digital signal processing filters makes a similar distinction and provides a non-streaming semantics for reasoning about programs and a streaming semantics for executing programs.

@@

Rule {\sc Prim} states that to evaluate a primitive $p$ applied to many arguments $e_1$ to $e_n$, we evaluate each argument separately; we then use the prim-sem metafunction to apply the primitive.

@@

Rule $\textsc{Fby}_S$ evaluates a followed-by expression when the streaming history contains multiple entries.
In this case, $\xfby{v}{e}$ proceeds to evaluate the previous value of $e$ by discarding the most recent entry from the streaming history.

Rule {\sc Rec} evaluates a recursive stream $\xrec{x}{e}$ by unfolding the recursion one step.
For causal expressions (\autoref{s:core:causality}), where each recursive occurrence of $x$ is guarded by a followed-by, this unfolding will eventually terminate, as the follow-by shortens the streaming history.

@@

We also use two auxiliary judgment forms: $\bigsteps{\Sigma}{e}{V}$ and $\bigstepalways{\Sigma}{e}$.

Judgment form $\bigsteps{\Sigma}{e}{V}$ denotes that, under streaming history $\Sigma$, expression $e$ evaluates to the \emph{stream} $V$.
This judgment is an iterated application of the single-value big-step form.

Judgment form $\bigstepalways{\Sigma}{e}$ denotes that expression $e$, which must be a boolean, evaluates to the stream of trues under history $\Sigma$.
Informally, it can be read as ``in streaming history $\Sigma$, $e$ is always true''.

@@

To prove that a contract \emph{instantiation} (a call-site) is valid, we show that, under the calling environment, the rely clause is always true.
Crucially, the proof can also use the fact that, if the rely is always true, then the guarantee is always true.
This sort of feedback is necessary for proving properties of mutually-dependent calls.
Although this feedback appears circular, we enforce causality by requiring that occurrences of recursive streams are guarded by delays (\autoref{s:core:causality}).

We define the checked semantics of Pipit in \autoref{f:core-check}.
The checked semantics mostly follows the structure of the dynamic semantics, checking any properties and contracts as they are encountered.

@@

Rule {\sc ChkContract} applies when checking property status $\PStatus$ of a contract with expression $\xcontractP{\PStatus'}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$.
Although we only include one property status on the contract, conceptually there are two distinct properties: one for the caller ($\PStatus'$) and one for the definition itself (assumed to be $\PSValid$).
To check the caller property when $\PStatus = \PStatus'$, we evaluate the rely $\erely$ and require it to be true.
To check the definition property when $\PStatus = \PSValid$, we assume that the rely holds, and check that the body satisfies the guarantee.
We also descend into the subexpressions to check them; when checking the body and guarantee, we can assume that the rely holds.
Unfortunately, this rule must deal with the two different roles of a contract at once; in the next section, we will separate the two roles.

@@

Blessing is slightly different for contract definitions, as we need to separate the definition of the contract from the instantiation.
To check that a contract definition is valid, we show that if the rely clause is always true for a particular input, then the body satisfies the guarantee for the same inputs.
We also assume that the valid properties in the rely, body and guarantee hold, and show the corresponding unknown properties:

@@

\begin{theorem}[bigstep-substitute]
  For a streaming history $\Sigma$ and a causal expression $e$, if $e[x := e']$ evaluates to a value $v$ $(\bigstep{\Sigma}{e}{v})$, then we can evaluate $e'$ to some stream $V$ and extend the streaming history to evaluate $e$ to the original value $(\bigstep{\Sigma[x \mapsto V]}{e}{v})$.
  The converse is also true.
\end{theorem}

@@

The semantics in \autoref{f:core-bigstep} for a recursive expression $\xrec{x}{e}$ performs one step of recursion by substituting $x$ for the recursive expression.
An alternative semantics would be to have the environment outside the semantics invent a stream $V$ such that if we extend the streaming history with $x \mapsto V$, then $e$ evaluates to $V$ itself.
The above substitution theorem can be used to show that these two semantics are equivalent.
Thanks to causality, we can additionally show that, when evaluating $e$ with $x \mapsto V$, the most recent value in $V$ does not affect the result.
This fact can be used to ``seed'' evaluation by starting with an arbitrary value:
\begin{theorem}[bigstep-rec-causal]
  For a streaming history $\Sigma; \sigma$ and a causal recursive expression $\xrec{x}{e}$, if $(\bigstep{\Sigma; \sigma}{e}{v})$, then updating $\sigma[x]$ with any value $v'$ results in the same value: $(\bigstep{\Sigma; \sigma[x \mapsto v']}{e}{v})$.
\end{theorem}

>>>>

\autoref{f:core:structure} shows the high-level architecture of Pipit.
On the left-hand-side, the surface syntax embedded in \fstar{} is shown; this includes some Pipit-specific syntactic sugar.
The translation from the surface syntax to the core language is trusted.
There are two targets from the core language: abstract transition systems for verification, and executable transition systems for extraction to C.
The translation to abstract systems is verified to be an abstraction according to the dynamic semantics (\autoref{s:core:dynamic}).
One property that still remains to be proven is that the proof obligations on the abstract transition system entail the original proof obligations (\autoref{s:transition}); this pending proof is denoted as negated entailment in the figure ($\not\vdash$).
The translation to executable transition system is proven to be semantics preserving, as is the subsequent translation to \lowstar{}.
The translation from \lowstar{} to C is external to this paper and forms part of our trusted computing base.

@@

Recursive streams, which can refer to previous values of the stream itself, are defined using the fixpoint operator ($\xrec{x}{e[x]}$); the syntax $e[x]$ means that the variable $x$ can occur in $e$.
% To ensure that streams are productive,
As in Lustre, recursive streams can only refer to their previous values and must be \emph{guarded} by a delay: the stream $(\xrec{x}{\xfby{0}{(x + 1)}})$ is well-defined, but stream $(\xrec{x}{x + 1})$ is invalid and has no computational interpretation.
This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
% We use this form to define a substitution-based operational semantics that is syntax-directed, as opposed to the mutually-recursive form in \cite{caspi1995functional} which is not syntax-directed.
% Our semantics has a simpler proof of determinism; we believe it has simplified other necessary proofs too and will perform further evaluation.
Although we cannot express mutually-recursive bindings in the core syntax here, we can express them as a notation on the surface syntax by combining the bindings together into a record or tuple.

@@

Contracts $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ involve two verification conditions.
Firstly, when a contract is \emph{defined}, the definer must prove that the body satisfies the contract: roughly, if $\erely$ is always true, then $\eguar[x := \ebody]$ is always true.
Secondly, when a contract is \emph{instantiated}, the caller must prove that the environment satisfies the precondition: that is, $\erely$ is always true.
Conceptually, then, a contract could have two property statuses: one for the definition and one for the instantiation.
However, in practice, it is not useful to defer the proof of a contract definition --- one could achieve a similar effect by replacing the contract with its implementation.
For this reason, we only annotate contracts with one property status, which denotes whether the instantiation has been proved to satisfy the precondition.

@@

For values, we use an auxiliary definition $\mtypingval{v}{\tau}$ to denote that value $v$ has type $\tau$.
Likewise, for primitives we use $\mtypingprim{p}{(\tau_1 \times \cdots \hdots \times \tau_n) \to \tau'}$ to denote that primitive $p$ takes arguments of type $\tau_i$ and returns a result of type $\tau'$.
Primitives are pure, non-streaming functions.

@@

Finally, rule \textsc{TContract} applies for a contract $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ with a body expression of some type $\tau$.
The overall expression has result type $\tau$.
Both rely and guarantee clauses must be boolean expressions.
Additionally, the guarantee clause can refer to the result value as $x$.

@@

The dynamic semantics of Pipit are defined in \autoref{f:core-bigstep}.
We present our semantics in a big-step form.
This differs somewhat from traditional \emph{reactive} semantics of Lustre~\cite{caspi1995functional}.
Our big-step semantics emphasises the equational nature of Pipit, as it is substitution-based and syntax-directed, while the reactive semantics emphasises the finite-state streaming execution of the system.
We use transition systems for reasoning about the finite-state execution (\autoref{s:transition}), which is fairly standard~\cite{brun2023equation,champion2016kind2,raymond2008synchronous}.
Previous work on the {\sc W-calculus}~\cite{gallego2021w} for linear digital-signal-processing filters makes a similar distinction and provides a non-streaming semantics for reasoning about programs and a streaming semantics for executing programs.

@@

Rule {\sc Prim} states that to evaluate a primitive $p$ applied to many arguments $e_1$ to $e_n$, we evaluate each argument separately; we then apply the primitive with prim-sem metafunction.

@@

Rule $\textsc{Fby}_S$ evaluates a followed-by expression when the streaming history contains multiple entries.
In this case, $\xfby{v}{e}$ evaluates the previous value of $e$ by discarding the most recent entry from the streaming history.

Rule {\sc Rec} evaluates a recursive stream $\xrec{x}{e}$ by unfolding the recursion one step.
For causal expressions (\autoref{s:core:causality}), where each recursive occurrence of $x$ is guarded by a followed-by, this unfolding eventually terminates as each followed-by shortens the history.

@@

We also define two auxiliary judgment forms: $\bigsteps{\Sigma}{e}{V}$ and $\bigstepalways{\Sigma}{e}$.

Judgment form $\bigsteps{\Sigma}{e}{V}$ denotes that, under history $\Sigma$, expression $e$ evaluates to the \emph{stream} $V$.
This judgment performs iterated application of single-value evaluation.

Judgment form $\bigstepalways{\Sigma}{e}$ denotes that a boolean expression $e$ evaluates to the stream of trues under history $\Sigma$.
Informally, it can be read as ``$e$ is always true in history $\Sigma$''.

@@

To prove that a contract \emph{instantiation} (a call-site) is valid, we show that, under the calling environment, the rely clause is always true.
Crucially, the proof can also use the fact that, if the rely is always true, then the guarantee is always true.
This sort of feedback is necessary for proving properties of mutually-dependent calls.
This circular dependency is well-founded as our causality check ensures that recursive streams are guarded by delays (\autoref{s:core:causality}).

We define the checked semantics of Pipit in \autoref{f:core-check}.
The checked semantics mostly follows the structure of the dynamic semantics, additionally checking any properties and contracts as they are encountered.

@@

Rule {\sc ChkContract} applies when checking property status $\PStatus$ of a contract with expression $\xcontractP{\PStatus'}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$.
Although we only include one property status on the contract, conceptually there are two distinct properties: one for the caller ($\PStatus'$) and one for the definition itself (assumed to be $\PSValid$).
To check the caller property when $\PStatus = \PStatus'$, we evaluate the rely $\erely$ and require it to be true.
To check the definition property when $\PStatus = \PSValid$, we assume that the rely holds, and check that the body satisfies the guarantee.
We also descend into the subexpressions to check them; when checking the body and guarantee, we can assume that the rely holds.
% This rule must deal with the two different roles of a contract at once; in the next section, we will separate the two roles.

@@

Blessing is different for contract definitions, as we need to separate the definition of the contract from the instantiation.
To check that a contract definition is valid, we show that if the rely clause is always true for a particular input, then the body satisfies the guarantee for the same inputs.
We also assume that the valid properties in the rely, body and guarantee hold, and show the corresponding unknown properties:

@@

\begin{theorem}[bigstep-substitute]
  For a streaming history $\Sigma$ and causal expressions $e$ and $e'$, if $e[x := e']$ evaluates to a value $v$ $(\bigstep{\Sigma}{e}{v})$, then we can evaluate $e'$ to some stream $V$ and extend the streaming history to evaluate $e$ to the original value $(\bigstep{\Sigma[x \mapsto V]}{e}{v})$.
  The converse is also true.
\end{theorem}

@@

The semantics in \autoref{f:core-bigstep} for a recursive expression $\xrec{x}{e}$ performs one step of recursion by substituting $x$ for the recursive expression.
An alternative non-syntax-directed semantics would be to have the environment outside the semantics supply a stream $V$ such that if we extend the streaming history with $x \mapsto V$, then $e$ evaluates to $V$ itself.
The above substitution theorem can be used to show that, for causal expressions, these two semantics are equivalent.
We can additionally show that, when evaluating $e$ with $x \mapsto V$, the most recent value in $V$ does not affect the result.
This fact can be used to ``seed'' evaluation by starting with an arbitrary value:
\begin{theorem}[bigstep-rec-causal]
  For a streaming history $\Sigma; \sigma$ and a causal recursive expression $\xrec{x}{e}$, if $(\bigstep{\Sigma; \sigma}{e}{v})$, then updating $\sigma[x]$ with any value $v'$ results in the same value: $(\bigstep{\Sigma; \sigma[x \mapsto v']}{e}{v})$.
\end{theorem}

### 2024-01-16 2024pipit/section/40-Transition.tex
copy

\autoref{f:system-types} shows the types of transition systems.
A transition system is parameterised by its input context and the result type.
It also contains two internal contexts: firstly, the state context describes the private state required to execute the machine; secondly, the free context contains any extra input values that the transition system would like to quantify over.
The free context is used to allow the system to ask for arbitrary values from the environment, when it would not otherwise be able to return a concrete value.

For contract instantiations, which abstract over the implementation, the natural translation to a transition system would involve an existential quantifier: there exists some value that satisfies the specification.
Unfortunately, such an existential quantifier requires a step \emph{relation} rather than a step \emph{function}.
Using a step relation complicates the resulting transition system, as other operations such as primitive application must also introduce existential quantifiers; such quantifiers block normalisation and result in a more complex transition system.
Instead, the free context provides the step function with a fresh unconstrained value of the desired type, which the step function can then constrain.

As usual, the step-result contains the updated state for the transition system, as well as the result value.
The step-result additionally contains two propositions for the `rely', or assumptions about the execution environment, and `guarantee', or obligations that the transition system must show.
For the transition system corresponding to an expression $e$, these propositions are analogous to the known checked semantics $\semcheck{\Sigma}{\PSValid}{e}$ and unknown checks $\semcheck{\Sigma}{\PSUnknown}{e}$ respectively.

@@

\autoref{f:system-translation-contexts} defines the internal state and free contexts required for an expression.
For most expression forms, the state and free contexts are defined by taking the union of the contexts of subexpressions.
Followed-by delays introduce a local state variable $x_{@fby@}$ in which to store the most recent stream value.
We generate a fresh variable here, though the implementation uses de Bruijn indices.
Recursive streams and contracts both introduce new bindings into the free context, assuming that their binders $x$ are unique.

@@

To translate a followed-by $\xfby{v}{e}$, we initialise the follow-by's unique binder $x_{@fby@}$ to $v$.
At each step, we return the value in the local state, before updating the local state to the subexpression's new value.
The rely and guarantee differ from the checked semantics here: in the checked semantics, we check the subexpression on the previous inputs, but here we check the current subexpression.
This means that a single step of the rely and guarantee do not exactly correspond to the checked semantics; however, we posit that they are equivalent for a rely and guarantee that has been proven to hold for any sequence of inputs.

@@

To translate let-expressions $\xlet{x}{e}{e'}$, we extend the input heap with the value of $e$ before evaluating $e'$.
The presentation here duplicates the computation of the value of $e$, but this is not an issue in practice.

@@

In the contract instantiation, we assume that if the contract rely is true \emph{at the current step}, then the contract guarantee also holds at the current step.
The true semantics of the contract, however, only holds if the contract rely is true \emph{at every step so far}.
This simplification is justified as our definition of validity for a transition system also requires the translation rely to be true at every step.
This simplification is essentially an application of the $\Box \Box p \implies \Box p$ axiom of modal logic.
The mechanised proof of this simplification is future work.

>>>>

\autoref{f:system-types} shows the types of transition systems.
A transition system is parameterised by its input context and the result type.
It also contains two internal contexts: firstly, the state context describes the private state required to execute the machine; secondly, the free context contains any extra input values that the transition system would like to existentially quantify over.
The free context is used to allow the system to ask for arbitrary values from the environment, when it would not otherwise be able to return a concrete value.

For contract instantiations, which abstract over the implementation, the natural translation to a transition system would involve an existential quantifier: ``there exists some value that satisfies the specification''.
Unfortunately, such an existential quantifier requires a step \emph{relation} rather than a step \emph{function}.
Using a step relation complicates the resulting transition system, as other operations such as primitive application must also introduce existential quantifiers; such quantifiers block simplifications such as partial-evaluation and result in a more complex transition system.
Instead, the free context provides the step function with a fresh unconstrained value of the desired type, which the step function can then constrain.

Back to \autoref{f:system-types}, the step-result contains the updated state for the transition system, as well as the result value.
The step-result additionally contains two propositions for the `rely', or assumptions about the execution environment, and `guarantee', or obligations that the transition system must show.
For the transition system corresponding to an expression $e$, these propositions are analogous to the known checked semantics $\semcheck{\Sigma}{\PSValid}{e}$ and unknown checks $\semcheck{\Sigma}{\PSUnknown}{e}$ respectively.

@@

\autoref{f:system-translation-contexts} defines the internal state and free contexts required for an expression.
For most expression forms, the state and free contexts are defined by taking the union of the contexts of subexpressions.
Followed-by delays introduce a local state variable $x_{@fby@(e)}$ in which to store the most recent stream value.
We generate a fresh variable here, though the implementation uses de Bruijn indices.
Recursive streams and contracts both introduce new bindings into the free context; we assume that their binders $x$ are unique.

@@

To translate a followed-by $\xfby{v}{e}$, we initialise the follow-by's unique binder $x_{@fby@(e)}$ to $v$.
At each step, we return the value in the local state, before updating the local state to the subexpression's new value.
The rely and guarantee differ from the checked semantics here: in the checked semantics, we check the subexpression on the previous inputs, but here we check the current subexpression.
This means that a single step of the rely and guarantee do not exactly correspond to the checked semantics; however, we posit that they are equivalent for a rely and guarantee that has been proven to hold for any sequence of inputs.

@@

To translate let-expressions $\xlet{x}{e}{e'}$, we extend the input heap with the value of $e$ before evaluating $e'$.
The presentation here duplicates the computation of the value of $e$, but the actual implementation introduces a single binding.

@@

In the contract instantiation, we assume that if the contract rely is true \emph{at the current step}, then the contract guarantee also holds at the current step.
The real semantics of the contract, however, requires the contract rely to be true \emph{at every step so far}.
This difference is benign, as the inductive case of the proof of transition system validity assumes that both the rely and guarantee held at previous steps; if the rely also holds now, then it has held at every step so far.
% to prove that a transition system satisfies its guarantee, we show that if the rely is true \emph{at every step so far}, then the guarantee is also true.
% This simplification is essentially an application of the $\Box \Box p \iff \Box p$ equivalence of modal logic.
% The mechanised proof of this simplification is future work.

### 2024-01-16 2024pipit/section/50-Extraction.tex
copy

This translation to transition systems is verified to preserve the original semantics.
To extract the program, we use a \emph{hybrid embedding} as described in \cite{ho2022noise}, which is similar to staged-compilation.
The hybrid embedding involves a deep embedding of the Pipit core language, while the translation to executable transition systems produces a shallow embedding.
We use the \fstar{} host language's normalisation-by-evaluation and tactic support~\cite{martinez2019meta} to specialise the application of the translation to a particular input program.
This specialisation results in a concrete transition system that fits in the \lowstar{}~\cite{protzenko2017verified} subset of \fstar{}, which can then be extracted to statically-allocated C code.

>>>>

This translation to transition systems is verified to preserve the original semantics.
The invariant is very similar to that of \autoref{s:transition:proof}, except that the invariant descends into the implementations of contracts.
For the abstract systems we only showed abstraction; to prove that executable systems are equivalent to the original semantics, we rely on the fact that the original semantics and transition systems are both deterministic and total (\autoref{s:core:causality}).

To extract the program, we use a \emph{hybrid embedding} as described in \cite{ho2022noise}, which is similar to staged-compilation.
The hybrid embedding involves a deep embedding of the Pipit core language, while the translation to executable transition systems produces a shallow embedding.
We use the \fstar{} host language's normalisation-by-evaluation and tactic support~\cite{martinez2019meta} to specialise the application of the translation to a particular input program.
This specialisation results in a concrete transition system that fits in the \lowstar{} subset of \fstar{}, which can then be extracted to statically-allocated C code~\cite{protzenko2017verified}.

### 2024-01-16 2024pipit/section/60-Evaluation.tex
copy

The enable-acks field denotes whether the hardware should respond to messages from other nodes with an acknowledgement bit; in the case of a severe error acknowledgements are disabled, as the node must not write to the bus at all.
The transmit fields denote whether to send a reference message or an application-specific message.
For application-specific messages, the hardware-interop layer maintains the transmission buffers containing the actual message payload.
To meet the schedule as closely as possible, the driver anticipates the next transmission and includes a transmission delay to tell the hardware exactly when to send the next message.

>>>>

The enable-acknowledgements field denotes whether the hardware should respond to messages from other nodes with an acknowledgement bit; in the case of a severe error acknowledgements are disabled, as the node must not write to the bus at all.
The transmit fields denote whether to send a reference message or an application-specific message.
For application-specific messages, the hardware-interop layer maintains the transmission buffers containing the actual message payload.
To meet the schedule as closely as possible, the driver anticipates the next transmission and includes a transmission delay to tell the hardware exactly when to send the next message.

### 2024-01-16 2024pipit/section/70-Related.tex
copy

Verifying the time-triggered CAN driver is trickier, as the restrictions placed on the triggers array --- that triggers are sorted by time-mark, there must be an adequate time-gap between a trigger and its next-enabled, and a trigger's time-mark must be greater-than-or-equal-to its index --- naturally require quantifiers.
As described in \autoref{s:evaluation}, the Kind2 does include experimental array and quantifier support, but is limited to verifying small arrays up to 8 or 16 triggers.
Additionally, due to the limitations on top-level array definitions, compiling the program with Lustre~V6 would result in the entire triggers array being copied to the stack each iteration.

@@

Recent work has also introduced a form of refinement types for Lustre~\cite{chen2022synchronous}.
Rather than using transition systems, this work generates self-contained verification conditions based on the types of streams.
Such a type-based approach promises to allow abstraction of the implementation details.
However, for general-purpose functions such as \emph{count_when} from \autoref{s:motivation}, it is not clear how to give it a specification that actually \emph{abstracts} the implementation: a simple specification that the result is within some range would hide be insufficient for verifying the rest of the system.
For this function, the best specification is likely to include a re-statement of the implementation itself.

@@

Early work embedding a denotational semantics of Lucid Synchrone in an interactive theorem prover focussed on the semantics itself, rather than proving programs~\cite{boulme2001clocked}.
There is ongoing work to construct a denotational semantics of Vélus for program verification~\cite{bourke2022towards}.
We believe that the hybrid SMT approach of \fstar{} will allow for a better mixture of automated proofs with manual proofs.
Compared to Vélus alone, the trusted computing base of Pipit is larger: we depend on all of \fstar{}, \lowstar{}'s C code extraction and the Z3 SMT solver.

>>>>

Verifying the time-triggered CAN driver is trickier, as the restrictions placed on the triggers array --- that triggers are sorted by time-mark, there must be an adequate time-gap between a trigger and its next-enabled, and a trigger's time-mark must be greater-than-or-equal-to its index --- naturally require quantifiers.
As described in \autoref{s:evaluation}, the Kind2 does include experimental array and quantifier support, but in our experiments was limited to verifying small arrays up to 8 or 16 triggers.
Additionally, due to the limitations on top-level array definitions, compiling the program with Lustre~V6 would result in the entire triggers array being copied to the stack each iteration.

@@

Recent work has also introduced a form of refinement types for Lustre~\cite{chen2022synchronous}.
Rather than using transition systems, this work generates self-contained verification conditions based on the types of streams.
Such a type-based approach promises to allow abstraction of the implementation details.
However, for general-purpose functions such as \emph{count_when} from \autoref{s:motivation}, it is not clear how to give it a specification that actually \emph{abstracts} the implementation: a simple specification that the result is within some range would hide too much and be insufficient for verifying the rest of the system.
For such functions, the best specification is likely to include a re-statement of the implementation itself.

@@

Early work embedding a denotational semantics of Lucid Synchrone in an interactive theorem prover focussed on the semantics itself, rather than proving programs~\cite{boulme2001clocked}.
There is ongoing work to construct a denotational semantics of Vélus for program verification~\cite{bourke2022towards}.
We believe that the hybrid SMT approach of \fstar{} will allow for a better mixture of automated proofs with manual proofs.
Compared to Vélus alone, the trusted computing base of Pipit is larger: we depend on all of \fstar{}, \lowstar{}'s unverified C code extraction and the Z3 SMT solver; in comparison, Vélus' C code generation is verified and does not depend on any SMT solver.

### 2024-01-16 2024pipit/section/80-Conclusion.tex
copy

In the future, we intend to verify the remainder of the TTCAN driver logic.
We also intend to increase the expressivity of Pipit by adding \emph{clocks}, which are used to describe partially-defined streams~\cite{caspi1995functional}.
Clocks are important for composing complex systems together and avoiding unnecessary computation; clocks may be useful if it becomes necessary to optimise the runtime of the TTCAN driver.

>>>>

In the future, we intend to verify the remainder of the TTCAN driver logic.
We also intend to increase the expressivity of Pipit by adding \emph{clocks}, which are used to describe partially-defined streams~\cite{caspi1995functional}.
Clocks are important for composing complex systems together and avoiding unnecessary computation; they may be useful if it becomes necessary to optimise the runtime of the TTCAN driver.
